\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
library(ggplot2)
library(dbplyr)
library(tidyverse)
library(lubridate)
library(Kendall)
library(tseries)
library(zoo)
library(knitr)
library(here)
here()

# Set your ggplot theme



# Load your datasets
elwha_stream <- read.csv(here("Final_Project/Data/Processed/Elwa_Stream_Wrangled.csv"))
elwha_stream$Date <- as.Date(elwha_stream$Date, format = "%Y-%m-%d")
#glimpse(elwha_stream)

elwha_sediment <- read.csv(here("Final_Project/Data/Processed/Elwa_Sediment_Wrangled.csv"))
elwha_sediment$Date <- as.Date(elwha_sediment$Date, format = "%Y-%m-%d")
#glimpse(Elwha_stream)

```


# Rationale and Research Questions



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 

### Exploratory Analysis: Sediment Loads Data
```{r include = FALSE}
sed_summary <- summary(elwha_sediment)

```

| Variable                  | Minimum      | Median   | Mean     | Maximum       |
|---------------------------|-----------|----------|----------|------------|
| Date                      | 2011-09-15 | 2014-03-24 | 2014-03-24 | 2016-09-30 |
| Discharge m<sup>3</sup>/s | 6.33      | 41.70    | 47.70    | 399.50     |
| Suspended Sediment Concentration (mg/L) | 0.227     | 169.500  | 945.363  | 13819.790  |
| Average Fine SS Fraction  | 0.4100    | 0.6850   | 0.6896   | 0.9940     |
| Total SS Load (tonnes)      | 0.1       | 634.4    | 7750.9   | 429806.7   |
| Fine SS Load (tonnes)           | 0.1       | 457.9    | 4533.7   | 269484.4   |
| Sand SS Load (tonnes)            | 0.00      | 186.70   | 3217.18  | 160322.30  |
| Ungauged Bedload (tonnes)               | -7710     | 139      | 1976     | 104444     |
| Total Sediment Discharge (tonnes) | 0.1       | 779.8    | 9886.9   | 466989.1   |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
exp_discharge <- ggplot(elwha_sediment, aes(x=Date, y = Discharge))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x = 'Date', 
       y = expression("Daily Discharge (" ~ m^3/s ~ ")"))+
  ggtitle('Daily Discharge over Time')
exp_discharge

dischargebydate.regression <- lm(data = elwha_sediment, Discharge ~ Date)
dischbydate_reg <- summary(dischargebydate.regression)

```



|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|-------------|
| **Intercept** | 96.818629  | 27.387137  | 3.535   | 0.000418 *** |
| **Date**      | -0.003041  | 0.001695   | -1.794  | 0.072918 .   |
| **R-squared** |            |            |         | 0.001746     |
| **Adj R-squared** |        |            |         | 0.001204     |

Discharge shows strong seasonal trends, with an insignificant slight decrease over time (p value = 0.072918). Seasonal trends appear to become stronger with time. Time series analysis will remove seasonal trends in the data and reveal changes before and after dam removal

### Exploratory Analysis: Streamgage Data
```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
summary_stream <- summary(elwha_stream)

```
| Variable          | Minimum       | Median     | Mean       | Maximum     |
|--------------------|------------|------------|------------|------------|
| Date              | 2011-10-17 | 2013-11-27 | 2013-12-11 | 2016-05-11 |
| Average Stage Height (m)  | 0.1608     | 0.9752     | 0.9908     | 2.6560     |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
exp_gage_ht <- ggplot(elwha_stream, aes(x=Date, y = Avg_Stage_Height))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(y = 'Average Stage Height (m)')+
  ggtitle('Average Stage Height Over Time')
exp_gage_ht

stagebydate.regression <- lm(data = elwha_stream, Avg_Stage_Height ~ Date)
stagebydate_summary<- summary(stagebydate.regression)
```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | -5.493e+00 | 2.753e-01  | -19.95  | <2e-16 ***    |
| **Date**      | 4.040e-04  | 1.715e-05  | 23.56   | <2e-16 ***    |
| **R-squared** |            |            |         | 0.2646        |
| **Adj R-squared** |        |            |         | 0.2641        |

Stage height shows strong seasonal trends, with a significant increase over time (p value <2e-16)
Time series analysis will remove seasonal trends and reveal changes before and after dam removal. 


\newpage

# Analysis


## Question 1: How Does Are Daily Water and Sediment Discharges Related?

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
totalsed_discharge <- ggplot(elwha_sediment, aes(x=Discharge, y = Suspended_sed_load_t))+
  geom_line()+
  geom_smooth(method = "lm")+
  labs(x = expression("Daily Discharge (" ~ m^3/s ~ ")"), 
       y = 'Total Suspended Sediment Load (tonnes)')+
  ggtitle('Total SS vs Discharge')

totalsed_discharge

dischargebydate.totalregression <- lm(data = elwha_sediment,Suspended_sed_load_t  ~ Discharge)
dischargebydate_totalsummary <- summary(dischargebydate.regression)
#significant relation, R2 = 0.56



```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | -16278.35  | 644.40     | -25.26  | <2e-16 ***    |
| **Discharge** | 503.73     | 10.49      | 48.03   | <2e-16 ***    |
| **R-squared** |            |            |         | 0.5561        |
| **Adj R-squared** |        |            |         | 0.5559        |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
fine_fraction_discharge <- ggplot(elwha_sediment, aes(x=Discharge, y = Avg_fraction_fines))+
  geom_line()+
  geom_smooth(method = "lm")+
  labs(x = expression("Daily Discharge (" ~ m^3/s ~ ")"), 
       y = 'Average Fine Sediment Fraction')+
  ggtitle('Fine SS Fraction vs Discharge')

fine_fraction_discharge

dischargebydate.finesregression <- lm(data = elwha_sediment, Avg_fraction_fines  ~ Discharge)
dischargebydate_finessummary <- summary(dischargebydate.regression)
#significant relation, R2 = 0.16


```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | 7.451e-01  | 3.693e-03  | 201.73  | <2e-16 ***    |
| **Discharge** | -1.165e-03 | 6.014e-05  | -19.36  | <2e-16 ***    |
| **R-squared** |            |            |         | 0.1699        |
| **Adj R-squared** |        |            |         | 0.1695        |

## Question 2: Time series?

## Question 3: Comparison of distributions?




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
