\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
library(ggplot2)
library(dbplyr)
library(tidyverse)
library(lubridate)
library(Kendall)
library(tseries)
library(zoo)
library(knitr)
library(here)
here()

# Set your ggplot theme



# Load your datasets
elwha_stream <- read.csv(here("Final_Project/Data/Processed/Elwa_Stream_Wrangled.csv"))
elwha_stream$Date <- as.Date(elwha_stream$Date, format = "%Y-%m-%d")
#glimpse(elwha_stream)

elwha_sediment <- read.csv(here("Final_Project/Data/Processed/Elwa_Sediment_Wrangled.csv"))
elwha_sediment$Date <- as.Date(elwha_sediment$Date, format = "%Y-%m-%d")
#glimpse(Elwha_stream)

```


# Rationale and Research Questions



\newpage

# Dataset Information

\newpage

# Data Wrangling
The datasets downloaded included many columns that were not useful for this analysis. For the sediment load dataset, the date column was reclassified as a date, then the useful columns were selected and renamed. The Date column in the Stream Gage dataset included a date and time, so those pieces were extracted into separate columns and the Date was reclassified. Measurements that were flagged as Bad, Buried, or Noisy were removed. Next, an average stage height measurement for each day was calculated. With these wrangling steps complete, the datasets are ready for initial visualizations. 

\newpage

# Exploratory Analysis 

### Exploratory Analysis: Sediment Loads Data
```{r include = FALSE}
sed_summary <- summary(elwha_sediment)

```

| Variable                  | Minimum      | Median   | Mean     | Maximum       |
|---------------------------|-----------|----------|----------|------------|
| Date                      | 2011-09-15 | 2014-03-24 | 2014-03-24 | 2016-09-30 |
| Discharge m<sup>3</sup>/s | 6.33      | 41.70    | 47.70    | 399.50     |
| Suspended Sediment Concentration (mg/L) | 0.227     | 169.500  | 945.363  | 13819.790  |
| Average Fine SS Fraction  | 0.4100    | 0.6850   | 0.6896   | 0.9940     |
| Total SS Load (tonnes)      | 0.1       | 634.4    | 7750.9   | 429806.7   |
| Fine SS Load (tonnes)           | 0.1       | 457.9    | 4533.7   | 269484.4   |
| Sand SS Load (tonnes)            | 0.00      | 186.70   | 3217.18  | 160322.30  |
| Ungauged Bedload (tonnes)               | -7710     | 139      | 1976     | 104444     |
| Total Sediment Discharge (tonnes) | 0.1       | 779.8    | 9886.9   | 466989.1   |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
exp_discharge <- ggplot(elwha_sediment, aes(x=Date, y = Discharge))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(x = 'Date', 
       y = expression("Daily Discharge (" ~ m^3/s ~ ")"))+
  ggtitle('Daily Discharge over Time')
exp_discharge

dischargebydate.regression <- lm(data = elwha_sediment, Discharge ~ Date)
dischbydate_reg <- summary(dischargebydate.regression)

```



|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|-------------|
| **Intercept** | 96.818629  | 27.387137  | 3.535   | 0.000418 *** |
| **Date**      | -0.003041  | 0.001695   | -1.794  | 0.072918 .   |
| **R-squared** |            |            |         | 0.001746     |
| **Adj R-squared** |        |            |         | 0.001204     |

Discharge shows strong seasonal trends, with an insignificant slight decrease over time (p value = 0.072918). Seasonal trends appear to become stronger with time. Time series analysis will remove seasonal trends in the data and reveal changes before and after dam removal

### Exploratory Analysis: Streamgage Data
```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
summary_stream <- summary(elwha_stream)

```
| Variable          | Minimum       | Median     | Mean       | Maximum     |
|--------------------|------------|------------|------------|------------|
| Date              | 2011-10-17 | 2013-11-27 | 2013-12-11 | 2016-05-11 |
| Average Stage Height (m)  | 0.1608     | 0.9752     | 0.9908     | 2.6560     |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
exp_gage_ht <- ggplot(elwha_stream, aes(x=Date, y = Avg_Stage_Height))+
  geom_point()+
  geom_smooth(method = "lm")+
  labs(y = 'Average Stage Height (m)')+
  ggtitle('Average Stage Height Over Time')
exp_gage_ht

stagebydate.regression <- lm(data = elwha_stream, Avg_Stage_Height ~ Date)
stagebydate_summary<- summary(stagebydate.regression)
```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | -5.493e+00 | 2.753e-01  | -19.95  | <2e-16 ***    |
| **Date**      | 4.040e-04  | 1.715e-05  | 23.56   | <2e-16 ***    |
| **R-squared** |            |            |         | 0.2646        |
| **Adj R-squared** |        |            |         | 0.2641        |

Stage height shows strong seasonal trends, with a significant increase over time (p value <2e-16)
Time series analysis will remove seasonal trends and reveal changes before and after dam removal. 


\newpage

# Analysis


## Question 1: How Does Are Daily Water and Sediment Discharges Related?

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
totalsed_discharge <- ggplot(elwha_sediment, aes(x=Discharge, y = Suspended_sed_load_t))+
  geom_line()+
  geom_smooth(method = "lm")+
  labs(x = expression("Daily Discharge (" ~ m^3/s ~ ")"), 
       y = 'Total Suspended Sediment Load (tonnes)')+
  ggtitle('Total SS vs Discharge')

totalsed_discharge

dischargebydate.totalregression <- lm(data = elwha_sediment,Suspended_sed_load_t  ~ Discharge)
dischargebydate_totalsummary <- summary(dischargebydate.regression)
#significant relation, R2 = 0.56



```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | -16278.35  | 644.40     | -25.26  | <2e-16 ***    |
| **Discharge** | 503.73     | 10.49      | 48.03   | <2e-16 ***    |
| **R-squared** |            |            |         | 0.5561        |
| **Adj R-squared** |        |            |         | 0.5559        |

```{r echo = FALSE, message = FALSE, warning = FALSE, error = FALSE}
fine_fraction_discharge <- ggplot(elwha_sediment, aes(x=Discharge, y = Avg_fraction_fines))+
  geom_line()+
  geom_smooth(method = "lm")+
  labs(x = expression("Daily Discharge (" ~ m^3/s ~ ")"), 
       y = 'Average Fine Sediment Fraction')+
  ggtitle('Fine SS Fraction vs Discharge')

fine_fraction_discharge

dischargebydate.finesregression <- lm(data = elwha_sediment, Avg_fraction_fines  ~ Discharge)
dischargebydate_finessummary <- summary(dischargebydate.regression)
#significant relation, R2 = 0.16


```

|               | Estimate   | Std. Error | t value | `Pr(>|t|)`    |
|---------------|------------|------------|---------|---------------|
| **Intercept** | 7.451e-01  | 3.693e-03  | 201.73  | <2e-16 ***    |
| **Discharge** | -1.165e-03 | 6.014e-05  | -19.36  | <2e-16 ***    |
| **R-squared** |            |            |         | 0.1699        |
| **Adj R-squared** |        |            |         | 0.1695        |

## Question 2: How did sediment discharge and river discharge change over time?

```{r, echo=FALSE}

Elwha_stream_daily <- elwha_stream 

Elwha_stream_early <- Elwha_stream_daily %>% 
  filter(Date < as.Date("2015-07-01"))
# generating time series for the average daily stream gage height 
Stream_daily.ts <- ts(Elwha_stream_early$Avg_Stage_Height, 
                      start = c(2011, 10), frequency = 365)
# decomposing the Time Series
stream_daily_decomp <- stl(Stream_daily.ts, s.window = 'periodic')
plot(stream_daily_decomp)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# removing the seasonal component 
nonseasonal_daily <- Stream_daily.ts -
stream_daily_decomp$time.series[,"seasonal"] 
#plot of non seasonal data 
plot(nonseasonal_daily)
# mann kendall test for non seasonal 
MannKendall(nonseasonal_daily)

```
For this time series, the seasonal component was removed in order to run a Mann Kendall test. The test produced a tau value of 0.388 and a p-value of < 2.22e-16. Since the p-value is less than 0.05 and the tau is positive, there is a statistically significant increase in stream gage height from 2011 to 2015. 
```{r, echo=FALSE}
discharge_daily.ts <- ts(elwha_sediment$Discharge, start = c(2011, 09), 
                         frequency = 365)
discharge_daily_decomp <- stl(discharge_daily.ts, s.window = 'periodic')
plot(discharge_daily_decomp)
SeasonalMannKendall(discharge_daily.ts)
```
A Seasonal Mann-Kendall test was performed on the data producing a tau value of 1 and a p-value of 2.22e-16. There is a statistically significant increasing trend for discharge over time. 

```{r, echo=FALSE}
sediment_discharge.ts <- ts(elwha_sediment$Total_Sed_discharge, start = c(2011, 09), frequency = 365)
sed_discharge_decomp <- stl(sediment_discharge.ts, s.window = 'periodic')
plot(sed_discharge_decomp)

SeasonalMannKendall(sediment_discharge.ts)
```
A seasonal Mann-Kendall test was performed on the data which produced a tau value of 0.69 and a p-value of < 2.22e-16. These results indicate a statistically significant increasing trend in total sediment discharge in the Elwha River during and after dam removal. 

## Question 3: Comparison of distributions?




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
