
Kayla's Working Rmd

To do:
- rationale (1-2 paragraphs with citations)
- Numbered list of research questions 
- Analysis #2: Comparison of sediment distributions before and after dam removal
(t-test before and after dam removal)
- contribute to conclusion



Libraries
```{r load libraries}
library(tidyverse)
library(here)
library(ggplot2)
library(lubridate)
```

Loading in Data
```{r loading in data}
#check working environment
here()

elwha_sediment_raw <- read_csv(here('Final_Project/Data/Raw/Elwha_DailySedimentLoads_2011to2016.csv'))


#read in wrangled stream data 
elwha_stream <- read_csv(here('Final_Project/Data/Processed/Elwa_Stream_Wrangled.csv'))
#fix dates 
#elwha_stream$Date <- as.Date(elwha_stream, format = "%Y-%m-%d")

#view stream data
glimpse(elwha_stream)
head(elwha_stream)

#read in wrangled sediment data
elwha_sediment <- read_csv(here('Final_Project/Data/Processed/Elwa_Sediment_Wrangled.csv'))

#view sediment data
glimpse(elwha_sediment)
head(elwha_sediment)
```

Analysis #2 - comparison of suspended sediment before and after dam removal
daily suspended sediment load

Dam Removal went from September 2011 to Summer 2014 
Before: all dates before 9/2011 (doesn't exist)
After: 9/2014

Analysis #2 pt 2 - 
T-test with how the during process compares to the after 

(maybe) time series analysis of suspended sediments during dam removal process

T-test 
why? a t-test can be used to determine is there is a statistically significant difference
between two paired data sets ,, a paired or twi sample t-test
tests whether a change or intervention 
had a meaningful impact on a group of subjects by comparing measurements before and after

Need to decide which continuous variable to look at to best understand 
```{r t-test}
#take a look at data first 
ggplot(elwha_sediment, aes(x = Date, y = Total_Sed_discharge)) +
  geom_line()

#okay, keep this idea but for t-test to work we need two factors that we are comparing so we will pick two years
#based on this graph we will choose 2013 vs 2015 

#select data from only 2013 and 2015, add Year as a column
elwha_sediment_1315 <- elwha_sediment %>% 
  mutate(Year = year(Date)) %>%
  filter(Year == 2013 | Year == 2015)

#now view
ggplot(elwha_sediment_1315, aes(x = Total_Sed_discharge, color = as.factor(Year))) +
  geom_freqpoly()

#Continuous dependent variable: Sediment Discharge 
#Categorical variable with two levels (2012 and 2016)
twosample_test_13to15 <- t.test(elwha_sediment_1315$Total_Sed_discharge ~ elwha_sediment_1315$Year)
twosample_test_13to15

#significant 



#test other years 
#2012 to 2015
elwha_sediment_1215 <- elwha_sediment %>% 
  mutate(Year = year(Date)) %>%
  filter(Year == 2012 | Year == 2015)

#now view
ggplot(elwha_sediment_1215, aes(x = Total_Sed_discharge, color = as.factor(Year))) +
  geom_freqpoly()

#run t-test
twosample_test_12to15 <- t.test(elwha_sediment_1215$Total_Sed_discharge ~ elwha_sediment_1215$Year)
twosample_test_12to15
#not significant 


#2011 to 2016
elwha_sediment_1116 <- elwha_sediment %>% 
  mutate(Year = year(Date)) %>%
  filter(Year == 2011 | Year == 2016)

#now view
ggplot(elwha_sediment_1116, aes(x = Total_Sed_discharge, color = as.factor(Year))) +
  geom_freqpoly()

#run t-test
twosample_test_11to16 <- t.test(elwha_sediment_1116$Total_Sed_discharge ~ elwha_sediment_1116$Year)
twosample_test_11to16

#not significant?
```


Sources:
https://www.usgs.gov/centers/pcmsc/science/usgs-science-supporting-elwha-river-restoration-project
https://geonarrative.usgs.gov/elwhariverrestoration/

